Type cones of $\b{g}$-vector fans

### 2019-04-02 thirdPaper/draft/typeConeAssociahedra.tex

\subsection{Type cone}

Fix an essential complete simplicial fan~$\Fan$ in~$\R^n$. Let~$\b{G}$ the $N \times n$-matrix whose rows are the rays of~$\Fan$ and let~$\b{K}$ be a $(N-n) \times N$-matrix that spans the left kernel of~$\b{G}$ (\ie $\b{K}\b{G} = 0$). For any height vector~$\b{h} \in \R^N$, we define the polytope
\[
P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}.
\]
We say that $h$ is \defn{$\Fan$-admissible} if $P_\b{h}$ is a polytopal realization of~$\Fan$.
The following classical statement characterizes the $\Fan$-admissible height vectors.
It is a reformulation of regularity of triangulations of vector configurations, introduced in the theory of secondary polytopes~\cite{GelfandKapranovZelevinsky}, see also~\cite{DeLoeraRambauSantos}.
We present here a convenient formulation from~\cite[Lem.~2.1]{ChapotonFominZelevinsky}.

\begin{proposition}
\label{prop:characterizationPolytopalFan}
Then the following are equivalent for any height vector~$h \in \R^N$:
\begin{enumerate}
\item The fan~$\Fan$ is the normal fan of the polytope~$P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}$.
\item For any two adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ with~$\rays \ssm \{\ray\} = \rays' \ssm \{\ray'\}$,
\[
\alpha \, h_{\ray} + \alpha' \, h_{\ray'} + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, h_{\ray[s]} > 0,
\]
where
\[
\alpha \, \ray + \alpha' \, \ray' + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \ray[s] = 0
\]
is the unique (up to rescaling) linear dependence with~$\alpha, \alpha' > 0$ between the rays of~$\rays \cup \rays'$.
\end{enumerate}
\end{proposition}

@@

In this paper, we are interested in the set of all possible realizations of~$\Fan$. This was studied by P.~McMullen in~\cite{McMullen-typeCone}.

\begin{definition}
The \defn{type cone} of~$\Fan$ is the cone~$\typeCone(\Fan)$ of all $\Fan$-admissible height vectors~$\b{h}$:
\begin{align*}
\typeCone(\Fan) & \eqdef \set{\b{h} \in \R^N}{\Fan \text{ is the normal fan of } P_\b{h}} \\
& = \Bigset{\b{h} \in \R^N}{\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0 \; \begin{array}{l} \text{for any adjacent maximal} \\ \text{cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$} \end{array}}.
\end{align*}
\end{definition}

Note that the type cone is an open cone and contains a linearity subspace of dimension~$n$ (it is invariant by translation in~$\b{G} \R^n$). We could thus intersect~$\typeCone(\Fan)$ by the kernel of~$\b{G}$, or consider the projection~$\b{K}\typeCone(\Fan)$.

\begin{definition}
An \defn{extremal adjacent pair} of~$\Fan$ is a pair of adjacent maximal cones~$\{\R_{\ge0}\rays, \R_{\ge0}\rays'\}$ of~$\Fan$ such that the corresponding inequality $\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0$ in the definition of the type cone~$\typeCone(\Fan)$ actually defines a facet of~$\typeCone(\Fan)$.
\end{definition}

In other words, extremal adjacent pairs define the extremal rays of the polar of the type cone~$\typeCone(\Fan)$.
Understanding the extremal adjacent pairs of~$\Fan$ enables to describe its type cone~$\typeCone(\Fan)$ and thus all its polytopal realizations.

\begin{remark}
\label{rem:dimTypeCone}
Since the type cone is an open~$N$ dimensional cone with a linearity subspace of dimension~$n$, it has at least~$N-n$ facets.
We say that the type cone is \defn{simplicial} when it has precisely $N-n$ facets.
\end{remark}

@@

\cref{lem:MinkowskiSum} ensures that convex combinations in the type cone correspond to Minkowski combinations of polytopes. This provides natural Minkowski summands for all polytopal realizations of~$\Fan$.

\begin{corollary}
Any polytope in~$\typeCone(\Fan)$ is the Minkowski sum of at most~$N-n$ polytopes corresponding to the rays of~$\b{K}\typeCone(\Fan)$.
\end{corollary}

@@

%sage: Q = GentleQuiver({'e1':{'e2':['a']}, 'e2':{'e3':['b']}}, [('a','b')])
%
%sage: Q.non_kissing_fan().rays()
%N( 1,  0,  0),
%N( 0,  0,  1),
%N( 1, -1,  0),
%N( 0,  0, -1),
%N( 0, -1,  0),
%N(-1,  0,  0),
%N( 0,  1,  0),
%N( 0,  1, -1)
%in 3-d lattice N
%
%sage: TCNKC = type_cone_NKC(Q)
%sage: TCNKC
%A 8-dimensional polyhedron in QQ^8 defined as the convex hull of 1 vertex, 5 rays, 3 lines
%sage: TCNKC.lines()
%(A line in the direction (0, 0, 1, 0, 1, 0, -1, -1),
% A line in the direction (1, 0, 0, 0, -1, -1, 1, 1),
% A line in the direction (0, 1, 0, -1, 0, 0, 0, -1))
%sage: TCNKC.inequalities()
%(An inequality (-1, 0, 1, -1, 0, 0, 0, 1) x + 0 >= 0,
% An inequality (1, 0, -1, 0, 1, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 1, 0, 0, 0, 0, -1, 1) x + 0 >= 0,
% An inequality (0, 0, 1, 0, -1, 1, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 0, 1, 0, 0, 1, -1) x + 0 >= 0)

@@

%sage: Q = GentleQuiver({'e1':{'e2':['a']}, 'e2':{'e3':['b']}, 'e3':{'e1':['c']}}, [('b','c')])
%
%sage: Q.non_kissing_fan().rays()
%N( 1,  0,  0),0
%N( 0,  0,  1),1
%N( 1,  0, -1),2
%N( 1, -1,  0),3
%N( 0,  0, -1),4
%N(-1,  0,  1),5
%N( 0, -1,  0),6
%N( 0, -1,  1),7
%N(-1,  0,  0),8
%N( 0,  1,  0),9
%N( 0,  1, -1)10
%in 3-d lattice N
%
%sage: TCNKC = type_cone_NKC(Q)
%sage: TCNKC
%A 11-dimensional polyhedron in QQ^11 defined as the convex hull of 1 vertex, 9 rays, 3 lines
%sage: TCNKC.lines()
%(A line in the direction (0, 0, 0, 1, 0, 0, 1, 1, 0, -1, -1),
% A line in the direction (1, 0, 1, 1, 0, -1, 0, 0, -1, 0, 0),
% A line in the direction (1, 1, 0, 1, -1, 0, 0, 1, -1, 0, -1))
%sage: TCNKC.inequalities()
%(An inequality (0, 0, 0, 1, 0, 1, 0, -1, 0, 0, 0) x + 0 >= 0,
% An inequality (-1, 0, 1, 0, 0, 0, 0, 0, 0, 1, -1) x + 0 >= 0,
% An inequality (1, 0, -1, 0, 1, 0, 0, 0, 0, 0, 0) x + 0 >= 0,
% An inequality (1, -1, 0, -1, 0, 0, 0, 1, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 0, -1, 1, 0, 0, 0, 0, 0, 0, 1) x + 0 >= 0,
% An inequality (0, 1, 0, 0, 0, 0, 1, -1, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 0, 0, 0, -1, -1, 1, 1, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 1, -1, -1, 0, 1, 0, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 0, 0, 0, 1, 0, 0, -1, -1, 1) x + 0 >= 0)
\end{example}

>>>>

\subsection{Type-cone}
\arnau{McMullen uses a dash between type and cone. Should we do it too?}
Fix an essential complete simplicial fan~$\Fan$ in~$\R^n$. Let~$\b{G}$ the $N \times n$-matrix whose rows are the rays of~$\Fan$ and let~$\b{K}$ be a $(N-n) \times N$-matrix that spans the left kernel of~$\b{G}$ (\ie $\b{K}\b{G} = 0$). For any height vector~$\b{h} \in \R^N$, we define the polytope
\[
P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}.
\]
We say that $h$ is \defn{$\Fan$-admissible} if $P_\b{h}$ is a polytopal realization of~$\Fan$.
The following classical statement characterizes the $\Fan$-admissible height vectors.
It is a reformulation of regularity of triangulations of vector configurations, introduced in the theory of secondary polytopes~\cite{GelfandKapranovZelevinsky}, see also~\cite{DeLoeraRambauSantos}.
We present here a convenient formulation from~\cite[Lem.~2.1]{ChapotonFominZelevinsky}.

\begin{proposition}
\label{prop:characterizationPolytopalFan}
Let~$\Fan$ be an essential complete simplicial fan in~$\R^n$. Then the following are equivalent for any height vector~$h \in \R^N$:
\begin{enumerate}
\item The fan~$\Fan$ is the normal fan of the polytope~$P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}$.
\item For any two adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ with~$\rays \ssm \{\ray\} = \rays' \ssm \{\ray'\}$,
\[
\alpha \, h_{\ray} + \alpha' \, h_{\ray'} + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, h_{\ray[s]} > 0,
\]
where
\[
\alpha \, \ray + \alpha' \, \ray' + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \ray[s] = 0
\]
is the unique (up to rescaling) linear dependence with~$\alpha, \alpha' > 0$ between the rays of~$\rays \cup \rays'$.
\end{enumerate}
\end{proposition}

@@

In this paper, we are interested in the set of all possible realizations of~$\Fan$ as the normal fan of a polytope~$P_\b{h}$. This was studied by P.~McMullen in~\cite{McMullen-typeCone} (see~\cite[Sect.~9.5]{DeLoeraRambauSantos} for a formulation in terms of chambers of triangulations of vector configurations).

\begin{definition}\label{def:typecone}
The \defn{type-cone} of~$\Fan$ is the cone~$\typeCone(\Fan)$ of all $\Fan$-admissible height vectors~$\b{h}$:
\begin{align*}
\typeCone(\Fan) & \eqdef \set{\b{h} \in \R^N}{\Fan \text{ is the normal fan of } P_\b{h}} \\
& = \Bigset{\b{h} \in \R^N}{\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0 \; \begin{array}{l} \text{for any adjacent maximal} \\ \text{cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$} \end{array}}.
\end{align*}
\end{definition}

Note that the type-cone is an open cone and contains a linearity subspace of dimension~$n$ (it is invariant by translation in~$\b{G} \R^n$). We could thus intersect~$\typeCone(\Fan)$ by the kernel of~$\b{G}$, or consider the projection~$\b{K}\typeCone(\Fan)$.

We denote by $\ctypeCone(\Fan)$ the closure of $\typeCone(\Fan)$. It is a closed polyhedral cone defined by the inequalities $\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} \geq 0$ for adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$.

\begin{remark}
 \label{rmk:non-simplicial}
Although we will only treat with simplicial fans, for completeness, note that \cref{prop:characterizationPolytopalFan} and \cref{def:typecone} can be easily adapted to non-simplicial fans. If~$\Fan$ is an arbitrary essential complete fan, it suffices to consider consider any simplicial refinement~$\Fan'$, and set some of the strict inequalities of the definition of $\typeCone(\Fan')$ to equality. Specifically, for any adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan'$ we impose
\[
 \sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} {\left\{\begin{array}{l} =\\ >\end{array}\right\}} 0
\]
with `$>$' when $\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ do not belong to the same maximal cone of $\Fan$; and with `$=$' if $\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ belong to the same maximal cone of $\Fan$.
\end{remark}

\begin{definition}
An \defn{extremal adjacent pair} of~$\Fan$ is a pair of adjacent maximal cones~$\{\R_{\ge0}\rays, \R_{\ge0}\rays'\}$ of~$\Fan$ such that the corresponding inequality $\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} \geq 0$ in the definition of the type-cone~$\ctypeCone(\Fan)$ actually defines a facet of~$\ctypeCone(\Fan)$.
\end{definition}
\arnau{I changed open to closed here}

In other words, extremal adjacent pairs define the extremal rays of the polar of the closed type-cone~$\ctypeCone(\Fan)$.
Understanding the extremal adjacent pairs of~$\Fan$ enables to describe its type-cone~$\typeCone(\Fan)$ and thus all its polytopal realizations.

\begin{remark}
\label{rem:dimTypeCone}
Since the type-cone is an open~$N$ dimensional cone with a linearity subspace of dimension~$n$, it has at least~$N-n$ facets.
We say that the type-cone is \defn{simplicial} when it has precisely $N-n$ facets.
\end{remark}

@@

We say that~$P$ is a \defn{Minkowski summand} of~$R$ if there is a polytope~$Q$ such that $R=P+Q$, and a \defn{weak Minkowski summand} if there is a $\lambda\geq 0$ and a polytope~$Q$ such that $\lambda R=P+Q$. 
\cref{lem:MinkowskiSum} ensures that convex combinations in the type-cone correspond to Minkowski combinations of polytopes. Actually, the set of weak Minkowski summands of a polytope $P$ with normal fan~$\Fan$ is isomorphic to the closed type-cone~$\ctypeCone(\Fan)$ modulo linearity (the cone of weak Minkowski summands is studied by Meyer in~\cite{Meyer1974}, to see that it is equivalent to the type-cone and other formulations, see for example the appendix of~\cite{PostnikovReinerWilliams}).

This provides natural Minkowski summands for all polytopal realizations of~$\Fan$: the rays of the type-cone. These correspond to indecomposable polytopes~(cf.~\cite{Meyer1974} or \cite{McMullen-typeCone}). A polytope $P$ in $\R^n$ is 
  called \defn{indecomposable} if all its weak Minkowski summands are of the form $\lambda P+t$ for some $\lambda\geq 0$ and $t\in\R^n$.

\begin{corollary}
\arnau{See \cite{Meyer1974} and \cite{McMullen-typeCone}}
Any polytope in~$\typeCone(\Fan)$ is the Minkowski sum of at most~$N-n$ indecomposable polytopes corresponding to the rays of~$\b{K}\typeCone(\Fan)$.
\end{corollary}

If $P=Q+R$, then we say that $R$ is the \defn{Minkowski difference} of $P$ and $Q$, which we denote by $R=P-Q$. We will only consider the case where $Q$ is a summand of $P$.
There are two natural ways this construction generalizes to arbitrary pairs of polyhedra. On the one hand, we have $P-Q=\set{x\in\R^n}{x+Q\subseteq P}$. 
Notice that $P_\b{h}-P_{\b{h}'}=P_{\b{h} - \b{h'}}$ when $P_{\b{h}'}$ is a Minkowski summand of $P_\b{h}$. Considering arbitrary differences of support functions gives rise to  \defn{virtual polyhedra}, which is the Grothendieck group of the semigroup of polytopes with Minkowski sums~\cite{PukhlikovKhovanskii1992}.

\begin{corollary}
Let $\b{h}_1,\dots, \b{h}_{N-n}\in \ctypeCone(\Fan)$ form a basis of $\ker \b{G}$. Then every $P_\b{h}$ with $\b{h}\in \ctypeCone(\Fan)$ has a unique representation (up to translation) as a signed Minkowski sum of dilates of $P_{\b{h}_1},\dots, P_{\b{h}_{N-n}}$.
\end{corollary}
\begin{proof}
 It suffices to express the projection of $\b{h}$ into $\ker\b{G}$ in terms of $\b{h}_1,\dots, \b{h}_{N-n}$, and to separate those terms with positive and negative coefficients to express Minkowski sums and differences.
\end{proof}

This is used in~\cite{ArdilaBenedettiDoker} to show that every generalized permutahedron can be written uniquelyas a signed Minkowski sum of simplices. 
\arnau{This has consequences in expressing volume and other valuations! I have to have a look at it.}
In the particular case when the type-cone is simplicial, then we can get rid of the Minkowski differences.
\begin{corollary}
 If the type-cone~$\typeCone(\Fan)$ of $\Fan$ is simplicial, then the $N-n$ idecomposable polytopes $P_{\b{h}_1},\dots, P_{\b{h}_{N-n}}$ arising from the rays of $\ctypeCone(\Fan)$
 verify that 
 \begin{enumerate}[(i)]
  \item  every $P_\b{h}$ with $\b{h}\in \typeCone(\Fan)$ has a unique representation (up to translation) as a Minkowski sum of positive dilates of $P_{\b{h}_1},\dots, P_{\b{h}_{N-n}}$; and
 \item  every $P_\b{h}$ with $\b{h}\in \ctypeCone(\Fan)$ has a unique representation (up to translation) as a Minkowski sum of dilates of $P_{\b{h}_1},\dots, P_{\b{h}_{N-n}}$.
 \end{enumerate}

\end{corollary}
 Note that such a family of polytopes giving rise to unique positive representations only exists when the type-cone is simplicial.

\arnau{By \cite{McMullen-Valuations} valuations on the polytope $\alpha_1P_{\b{h}_1}+\cdots+\alpha_{N-n}P_{\b{h}_{N-n}}$ (such as volume) are a polynomial on $\alpha_1\dots\alpha_{N-n}$ with coefficients the mixed volumes. This also works when the $\alpha$'s are not necessarily positive \cite{ArdilaBenedettiDoker}. Does this help in any way?}

@@

%sage: Q = GentleQuiver({'e1':{'e2':['a']}, 'e2':{'e3':['b']}}, [('a','b')])
%
%sage: Q.non_kissing_fan().rays()
%N( 1,  0,  0),
%N( 0,  0,  1),
%N( 1, -1,  0),
%N( 0,  0, -1),
%N( 0, -1,  0),
%N(-1,  0,  0),
%N( 0,  1,  0),
%N( 0,  1, -1)
%in 3-d lattice N
%
%sage: TCNKC = type_cone_NKC(Q)
%sage: TCNKC
%A 8-dimensional polyhedron in QQ^8 defined as the convex hull of 1 vertex, 5 rays, 3 lines
%sage: TCNKC.lines()
%(A line in the direction (0, 0, 1, 0, 1, 0, -1, -1),
% A line in the direction (1, 0, 0, 0, -1, -1, 1, 1),
% A line in the direction (0, 1, 0, -1, 0, 0, 0, -1))
%sage: TCNKC.inequalities()
%(An inequality (-1, 0, 1, -1, 0, 0, 0, 1) x + 0 >= 0,
% An inequality (1, 0, -1, 0, 1, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 1, 0, 0, 0, 0, -1, 1) x + 0 >= 0,
% An inequality (0, 0, 1, 0, -1, 1, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 0, 1, 0, 0, 1, -1) x + 0 >= 0)
%sage: TCNKC.lines()
%(A line in the direction (0, 0, 1, 0, 1, 0, -1, -1),
% A line in the direction (1, 0, 0, 0, -1, -1, 1, 1),
% A line in the direction (0, 1, 0, -1, 0, 0, 0, -1))

@@

%sage: Q = GentleQuiver({'e1':{'e2':['a']}, 'e2':{'e3':['b']}, 'e3':{'e1':['c']}}, [('b','c')])
%
%sage: Q.non_kissing_fan().rays()
%N( 1,  0,  0),0
%N( 0,  0,  1),1
%N( 1,  0, -1),2
%N( 1, -1,  0),3
%N( 0,  0, -1),4
%N(-1,  0,  1),5
%N( 0, -1,  0),6
%N( 0, -1,  1),7
%N(-1,  0,  0),8
%N( 0,  1,  0),9
%N( 0,  1, -1)10
%in 3-d lattice N
%
%sage: TCNKC = type_cone_NKC(Q)
%sage: TCNKC
%A 11-dimensional polyhedron in QQ^11 defined as the convex hull of 1 vertex, 9 rays, 3 lines
%sage: TCNKC.lines()
%(A line in the direction (0, 0, 0, 1, 0, 0, 1, 1, 0, -1, -1),
% A line in the direction (1, 0, 1, 1, 0, -1, 0, 0, -1, 0, 0),
% A line in the direction (1, 1, 0, 1, -1, 0, 0, 1, -1, 0, -1))
%sage: TCNKC.inequalities()
%(An inequality (0, 0, 0, 1, 0, 1, 0, -1, 0, 0, 0) x + 0 >= 0,
% An inequality (-1, 0, 1, 0, 0, 0, 0, 0, 0, 1, -1) x + 0 >= 0,
% An inequality (1, 0, -1, 0, 1, 0, 0, 0, 0, 0, 0) x + 0 >= 0,
% An inequality (1, -1, 0, -1, 0, 0, 0, 1, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 0, -1, 1, 0, 0, 0, 0, 0, 0, 1) x + 0 >= 0,
% An inequality (0, 1, 0, 0, 0, 0, 1, -1, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 0, 0, 0, -1, -1, 1, 1, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 1, -1, -1, 0, 1, 0, 0, 0, 0) x + 0 >= 0,
% An inequality (0, 0, 0, 0, 0, 1, 0, 0, -1, -1, 1) x + 0 >= 0)
%sage: TCNKC.lines()
%(A line in the direction (0, 0, 0, 1, 0, 0, 1, 1, 0, -1, -1),
% A line in the direction (1, 0, 1, 1, 0, -1, 0, 0, -1, 0, 0),
% A line in the direction (1, 1, 0, 1, -1, 0, 0, 1, -1, 0, -1))
\vincent{Todo}
\end{example}
